---
title: "Emerging species - Adaptive capacity"
author: "LGCarlson"
date: "9/29/2021"
output: 
  html_document:
    toc: true
    toc_float: false
    theme: flatly
    highlight: haddock
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}

library(tidyverse)
library(here)
library(janitor)

knitr::opts_chunk$set(echo = TRUE)

nth_element <- function(vector, starting_position, n) { 
  vector[seq(starting_position, length(vector), n)] 
}
```


### Read in social data

```{r read data,message=F,warning=F}
# full, raw, social indicator dataset
social_indicators_original<-read_csv(here::here("input_data", "socialIndicatorData_2011-2015_allUSA.csv")) 

# data to match COCA and Colburn Community names
match_names<-read_csv(here::here("metadata", "match_communities_Colburn_to_COCA.csv"))

# sub-dataset for latitudinal sorting
lat_sort_df <- match_names %>% select(COCACommunity,lat_sort)
```

### Read in change data and wrangle community name 

```{r,message=F,warning=F}
# pull in full change dataset
community_weighted_changes<-read_csv(here::here("input_data","CommunityAggregatedWeightedChangesWithSocVulnDat_10252021.csv")) %>%  select(-X1) %>% 
  select(Community, ProjectionScenario, TotalChangeValue,TotalChangeVolume)

# clean community names to match with social data
community_weighted_cleannames<-community_weighted_changes %>% 
  mutate(COCACommunity = Community) %>% 
  mutate(Community = gsub("_CT", ",CT", Community)) %>% 
  mutate(Community = gsub("_MA", ",MA", Community)) %>% 
  mutate(Community = gsub("_MD", ",MD", Community)) %>% 
  mutate(Community = gsub("_ME", ",ME", Community)) %>% 
  mutate(Community = gsub("_NC", ",NC", Community)) %>% 
  mutate(Community = gsub("_NH", ",NH", Community)) %>% 
  mutate(Community = gsub("_NJ", ",NJ", Community)) %>% 
  mutate(Community = gsub("_NY", ",NY", Community)) %>% 
  mutate(Community = gsub("_RI", ",RI", Community)) %>% 
  mutate(Community = gsub("_VA", ",VA", Community)) %>% 
  separate(Community, into = c("community_name","state"),sep = ",") %>% 
  mutate(community_name = snakecase::to_title_case(community_name)) %>% 
  mutate(COCA_clean = paste(community_name, state, sep = ", ")) %>% 
  semi_join(match_names, by = "COCA_clean") 

# change "Cape may" to "Cape May"
community_weighted_cleannames$COCA_clean[community_weighted_cleannames$COCA_clean == "Cape may, NJ"] = "Cape May, NJ"

#length(unique(community_weighted_cleannames$COCA_clean))

# pull in weighted change dataset
species_weighted_changes<-read_csv(here::here("input_data","SpeciesCommunityCFDERSWeightedChanges_10182021.csv"))  %>%  select(-X1)


# clean community names to match with social data
species_weighted_cleannames<-species_weighted_changes %>% 
  mutate(COCACommunity = Community) %>% 
  mutate(Community = gsub("_CT", ",CT", Community)) %>% 
  mutate(Community = gsub("_MA", ",MA", Community)) %>% 
  mutate(Community = gsub("_MD", ",MD", Community)) %>% 
  mutate(Community = gsub("_ME", ",ME", Community)) %>% 
  mutate(Community = gsub("_NC", ",NC", Community)) %>% 
  mutate(Community = gsub("_NH", ",NH", Community)) %>% 
  mutate(Community = gsub("_NJ", ",NJ", Community)) %>% 
  mutate(Community = gsub("_NY", ",NY", Community)) %>% 
  mutate(Community = gsub("_RI", ",RI", Community)) %>% 
  mutate(Community = gsub("_VA", ",VA", Community)) %>% 
  mutate(Community = gsub(",MARSHFIELD_", "_MARSHFIELD_", Community)) %>% 
  separate(Community, into = c("community_name","state"),sep = ",") %>% 
  mutate(community_name = snakecase::to_title_case(community_name)) %>% 
  mutate(COCA_clean = paste(community_name, state, sep = ", ")) %>% 
  semi_join(match_names, by = "COCA_clean") 

#length(unique(species_weighted_cleannames$COCA_clean))  should = 144
```

### Create baseline dataset

```{r,message=F,warning=F}
changes_expanded<-species_weighted_cleannames %>% 
  mutate_if(is.character, str_replace_all, pattern = ".combo.b", replacement = "") %>% 
  mutate_if(is.character, str_replace_all, pattern = "Future_", replacement = "") %>% 
  separate(ProjectionScenario, into = c("TimeHorizon","ClimateScenario","EnsembleStat","ChangeMeasure"),sep = "_", remove = F) %>% 
  mutate(EnsembleStat = ifelse(EnsembleStat == "cold", "pct5th", EnsembleStat)) %>% 
  mutate(EnsembleStat = ifelse(EnsembleStat == "warm", "pct95th", EnsembleStat)) %>% 
  mutate(ChangeMeasure = ifelse(is.na(ChangeMeasure) == T, "raw", ChangeMeasure))

table(changes_expanded$ChangeMeasure)

baseline_vals<-changes_expanded %>% 
  filter(ProjectionScenario == "Baseline") %>%
  dplyr::select(COCA_clean, CommonName, ProjectionValue) %>% 
  filter(!is.na(ProjectionValue)) %>% 
  dplyr::rename("BaselineProjection" = "ProjectionValue") 

```



### Create a function to wrangle weighted changes data 

*Calculate the percent increase for each species based on the percent change between ProjectionValue in the baseline scenario and the ProjectionValue in the alternate scenarios*

```{r,message=F,warning=F}
wrangle<-function(dftomodify, ceiling, percentcutoff, ensemblestat){
  
  # dftomodify is an object of class "dataframe" or "tibble" that contains a minimally processed version of "SpeciesCommunityCFDERSWeightedChangesBothRCPs.csv" 
  # ceiling is a numeric value defining what the threshold for ProportionVolume should be >= to
  # percent cutoff is a numeric value defining what the shreshold for PercentChange should be >= to
  # ensemble stat is a vector of characters describing which ensemble scenarios should be included in the data should be either one or multiple of the following c("mean","pct5th","pct95th")
  
  dftomodify %>% 
    filter(ProjectionScenario != "Baseline") %>% 
    mutate(ProportionVolume = if_else(is.na(ProportionVolume), 0, ProportionVolume)) %>% #many ProportionVolumes with NA
    filter(ProportionVolume <= ceiling) %>% 
    filter(ChangeMeasure == "raw") %>% 
    mutate(TimeHorizon = as.numeric(TimeHorizon)) %>% 
    dplyr::rename("ScenarioProjection" = "ProjectionValue") %>% 
    dplyr::select(COCA_clean,CommonName,ProjectionScenario,TimeHorizon,ClimateScenario,
                  EnsembleStat,ChangeMeasure,Statistic,ScenarioProjection) %>% 
    right_join(baseline_vals, by = c("COCA_clean","CommonName")) %>% 
    filter(BaselineProjection < ScenarioProjection) %>% 
    mutate(PercentChange = ((ScenarioProjection - BaselineProjection)/BaselineProjection)*100) %>% 
    filter(PercentChange > percentcutoff) %>% 
    left_join(match_names, by="COCA_clean") %>% 
    filter(!is.na(COCA_clean)) %>% 
    group_by(COCA_clean,lat_sort, TimeHorizon, 
             ClimateScenario, EnsembleStat) %>% 
    summarise(n=n(),.groups="keep") %>% 
    mutate(ClimateScenarioLabs = ifelse(ClimateScenario == "rcp45","RCP 4.5","RCP 8.5")) %>% 
    mutate(TimeHorizonLabs = ifelse(TimeHorizon == "2055", "Mid-century", "Late century")) %>% 
    filter(EnsembleStat %in% ensemblestat)
}
```




### Create data for 0.1% Proportion volume ceiling, >100% increase (Mean ensemble)

```{r,message=F,warning=F}
mean_100_ceiling<-wrangle(changes_expanded, 0.001, 100, c("mean")) %>% 
  mutate(Ceiling = "0.1% proportion volume ceiling") %>% 
  mutate(PercentIncrease = ">100% increase") %>% 
  rename("EmergingSpecies" = "n") %>% 
  filter(ClimateScenario == "rcp85")
```



### Create a function to plot the emerging species data created by wrangle. 

*This function only works for visualizing on ensemble stat at a time.*

```{r,message=F,warning=F}
plot_emerging_singleensemblestat<-function(dftoplot,pointcolorvalue,ceilinglabel,ensemblelabel,percentincreaselabel){
  
  # dftoplot is an object of class "dataframe" or "tibble" which has been processed by the wrangle function
  # ylabsortdf is an object of class "dataframe" or "tibble" which has been processed by the latsort function
  # ceilinglabel is a string, should be enclosed with quotations, that describes whether a total volume ceiling has been imposed by wrangle
  # ensemble label is a string, should be enclosed with quotations, that describes which ensemble statistic is being visualised (could be "Mean ensemble", "5% ensemble", or "95% ensemble") *this function can only plot one ensemble stat at a time
  # percentincrease label is a string, should be enclosed with quotations, that describes the number of emerging species given the percent increase threshold imposed by wrangle (could be 50% or 100% increase)
  
  dftoplot %>% 
ggplot(aes(x=fct_reorder(COCA_clean,lat_sort,.desc = T), y=EmergingSpecies)) + geom_point(color = pointcolorvalue) +
  #facet_grid(rows=vars(fct_reorder(TimeHorizonLabs,TimeHorizon, .desc = F)), cols=vars(ClimateScenarioLabs)) +
  facet_grid(cols=vars(TimeHorizon)) +
  coord_flip() +
  theme(axis.text.y = element_text(size =7)) +
  labs(x="", y=percentincreaselabel, title=ceilinglabel, subtitle = ensemblelabel) 
}
```



### Fig. 1 Visualize raw emerging species data

*0.1% Proportion volume ceiling, >100% increase (All ensemble stats)*

```{r,fig.height=18,fig.width=10,echo=F,message=F,warning=F}

plot_emerging_singleensemblestat(mean_100_ceiling,"black","Number of emerging species","Mean ensemble, >100% increase, 0.1% proportion volume ceiling","Emerging species (n)") 

```




### Rescale data

```{r,message=F,warning=F}
# rescale each time horizon individually

#0 to 1 rescale
rescaled_2055<-mean_100_ceiling %>% 
  filter(TimeHorizon == "2055") 

rescaled_2055$EmergingSpecies = scales::rescale(rescaled_2055$EmergingSpecies, to=c(0,1))

#0 to 1 rescale
rescaled_2100<-mean_100_ceiling %>% 
  filter(TimeHorizon == "2100") 

rescaled_2100$EmergingSpecies = scales::rescale(rescaled_2100$EmergingSpecies, to=c(0,1))

# put time horizons back together
mean_100_ceiling_rescaled<-bind_rows(rescaled_2055,rescaled_2100) 

mean_100_ceiling_rescaled_renamed<-mean_100_ceiling_rescaled %>% 
  select(COCA_clean,TimeHorizon,ClimateScenario,EnsembleStat,EmergingSpecies,lat_sort) %>% 
  rename("EmergingSpecies_rescaled" = "EmergingSpecies")

```



### Fig. 2 Visualize rescaled emerging species data

*0.1% Proportion volume ceiling, >100% increase (All ensemble stats)*
*Number of emerging species rescaled 0 to 1*

```{r,fig.height=18,fig.width=10,echo=F,message=F,warning=F}

plot_emerging_singleensemblestat(mean_100_ceiling_rescaled,"black","Number of emerging species, rescaled 0 to 1","Mean ensemble, >100% increase, 0.1% proportion volume ceiling","Emerging species (n)") 

```


```{r,message=F,warning=F}
emerging_species_final<-mean_100_ceiling %>% 
  left_join(match_names, by = "COCA_clean") %>% 
  left_join(mean_100_ceiling_rescaled_renamed, by = c("COCA_clean","TimeHorizon","ClimateScenario","EnsembleStat")) %>% 
  select(COCACommunity,COCA_clean,TimeHorizon,ClimateScenario,EnsembleStat,EmergingSpecies,EmergingSpecies_rescaled)

```

### Save output

#### Data

```{r}
#write_csv(emerging_species_final,here::here("output_data","emerging_species_100ceiling_rescaled.csv"))
```


#### Figures

```{r}
#ggsave(filename = here::here("output_figures/emerging_species","emerging_species_100ceiling.pdf"), plot = last_plot(), height = 18,width = 10)

#ggsave(filename = here::here("output_figures/emerging_species","emerging_species_100ceiling_rescaled.pdf"), plot = last_plot(), height = 18,width = 10)
```

